The Current State of R Tools for Nonlinear Least Squares Modeling

### 2023-07-20 RefactoringNLS/revision202307/RJournal2023-15Rev1ResubmissionRJstyle/StateOfRToolsForNLS.tex
fixup yaml

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Current State of R Tools for Nonlinear Least Squares Modeling},
  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
  colorlinks=true,
  linkcolor={red},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

\title{The Current State of R Tools for Nonlinear Least Squares
Modeling}
\author{John C. Nash
\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
Bhattacharjee
\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
\date{2023-7-16}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

Our Google Summer of Code project ``Improvements to \texttt{nls()}''
considered the features and limitations of the R function \texttt{nls()}
with the aim of improving and rationalizing R tools for nonlinear
regression. The rich features of \texttt{nls()} are weakened by several
deficiencies and inconsistencies such as a lack of stabilization of the
Gauss-Newton solver. Further considerations are the usability and
maintainability of the code base that provides the functionality
\texttt{nls()} offers. Various packages, in particular \texttt{nlsr}
(John C Nash and Duncan Murdoch
(\protect\hyperlink{ref-nlsr2023manual}{2023})), provide alternative
capabilities. We consider the differences in goals, approaches, and
features of different tools for nonlinear least squares modeling in R.
Discussion of these matters is relevant to improving R generally as well
as its nonlinear estimation tools.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{weeddf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(tt, weed)}
\FunctionTok{plot}\NormalTok{(weeddf, }\AttributeTok{main=}\StringTok{"Hobbs weed infestation data"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# model formulas}
\NormalTok{frmu }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ b1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ b2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{b3 }\SpecialCharTok{*}\NormalTok{ tt))}
\NormalTok{frms }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}} \DecValTok{100}\SpecialCharTok{*}\NormalTok{ c1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \DecValTok{10} \SpecialCharTok{*}\NormalTok{ c2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1} \SpecialCharTok{*}\NormalTok{c3 }\SpecialCharTok{*}\NormalTok{ tt))}
\NormalTok{frmt }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt) }\SpecialCharTok{/}\NormalTok{ scal))}
\CommentTok{\#}
\CommentTok{\# Starting parameter sets}
\NormalTok{stu1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =}\DecValTok{1}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b3 =} \DecValTok{1}\NormalTok{)}
\NormalTok{sts1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1 =} \DecValTok{1}\NormalTok{, }\AttributeTok{c2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{c3 =} \DecValTok{1}\NormalTok{)}
\NormalTok{stt1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{1}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{1}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Logistic3U}
\NormalTok{hobbs.res  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x)\{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} residual}
  \CommentTok{\# This variant uses looping}
  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{length}\NormalTok{(x) }\SpecialCharTok{!=} \DecValTok{3}\NormalTok{) }\FunctionTok{stop}\NormalTok{(}\StringTok{"hobbs.res {-}{-} parameter vector n!=3"}\NormalTok{)}
\NormalTok{  y  }\OtherTok{\textless{}{-}}  \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{, }
           \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{  res  }\OtherTok{\textless{}{-}}\NormalTok{  x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt)) }\SpecialCharTok{{-}}\NormalTok{ y}
\NormalTok{\}}

\NormalTok{hobbs.jac  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x) \{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} Jacobian}
\NormalTok{  jj  }\OtherTok{\textless{}{-}}  \FunctionTok{matrix}\NormalTok{(}\FloatTok{0.0}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{3}\NormalTok{)}
\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{  yy  }\OtherTok{\textless{}{-}}  \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt)}
\NormalTok{  zz  }\OtherTok{\textless{}{-}}  \FloatTok{1.0} \SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ yy)}
\NormalTok{  jj[tt,}\DecValTok{1}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   zz}
\NormalTok{  jj[tt,}\DecValTok{2}\NormalTok{]   }\OtherTok{\textless{}{-}}   \SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ yy}
\NormalTok{  jj[tt,}\DecValTok{3}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ yy }\SpecialCharTok{*}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt}
  \FunctionTok{attr}\NormalTok{(jj, }\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}}\NormalTok{ jj}
\NormalTok{  jj}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nls(formula = frmu, start = stu1, data = weeddf) : 
##   singular gradient
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nls(formula = frms, start = sts1, data = weeddf) : 
##   singular gradient
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nls(formula = frmt, start = stt1, data = weeddf) : 
##   singular gradient
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(unls1p)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## unls1p  -- ss= 2.587277 :  b1 = 196.1863  b2 = 49.09164  b3 = 0.3135697; 17  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(snls1p)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## snls1p  -- ss= 2.587277 :  c1 = 1.961863  c2 = 4.909164  c3 = 3.135697; 15  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(tnls1p)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## tnls1p  -- ss= 2.587277 :  Asym = 196.1863  xmid = 12.4173  scal = 3.189083; 12  itns
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{print}\NormalTok{(unlx1) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## residual sumsquares =  2.5873  on  12 observations
##     after  19    Jacobian and  25 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## b1               196.186         11.31      17.35  3.167e-08  -4.859e-09        1011  
## b2               49.0916         1.688      29.08  3.284e-10  -3.099e-08      0.4605  
## b3               0.31357      0.006863      45.69  5.768e-12   2.305e-06     0.04714
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pshort}\NormalTok{(snlx1) }\CommentTok{\# a short{-}form output}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## snlx1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 34 res/ 23 jac
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pshort}\NormalTok{(tnlx1) }\CommentTok{\# alternatively print(tnlx1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## tnlx1  -- ss= 2.5873 :  Asym = 196.19  xmid = 12.417  scal = 3.1891; 36 res/ 27 jac
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pnls}\NormalTok{(unlm1)  }\CommentTok{\# Short form of output}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## unlm1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pnls}\NormalTok{(snlm1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## snlm1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 7  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pnls}\NormalTok{(tnlm1) }\CommentTok{\# short form to give sum of squares, else use summary(tnlm1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## tnlm1  -- ss= 9205.4 :  Asym = 35.532  xmid = 43376  scal = -2935.4; 39  itns
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(gslnls)}
\NormalTok{ugslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmu, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stu1))}
\FunctionTok{pnls}\NormalTok{(ugslnls1) }\CommentTok{\# to get sum of squares}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## ugslnls1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
\end{verbatim}

>>>>

% !TeX root = RJwrapper.tex
\title{The Current State of R Tools for Nonlinear Least Squares Modeling}
\author{by John C. Nash and Arkajyoti Bhattacharjee}

\maketitle

\abstract{%
Our Google Summer of Code project ``Improvements to \texttt{nls()}'' considered the features and limitations of the R function \texttt{nls()} with the aim of improving and rationalizing R tools for nonlinear regression. The rich features of \texttt{nls()} are weakened by several deficiencies and inconsistencies such as a lack of stabilization of the Gauss-Newton solver. Further considerations are the usability and maintainability of the code base that provides the functionality \texttt{nls()} offers. Various packages, in particular \texttt{nlsr} (John C Nash and Duncan Murdoch (\protect\hyperlink{ref-nlsr2023manual}{2023})), provide alternative capabilities. We consider the differences in goals, approaches, and features of different tools for nonlinear least squares modeling in R. Discussion of these matters is relevant to improving R generally as well as its nonlinear estimation tools.
}

@@

\begin{verbatim}
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12
weeddf <- data.frame(tt, weed)
plot(weeddf, main="Hobbs weed infestation data")
\end{verbatim}

@@

\begin{verbatim}
# model formulas
frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
frms <- weed ~ 100* c1 / (1 + 10 * c2 * exp(-0.1 *c3 * tt))
frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
#
# Starting parameter sets
stu1<-c(b1 =1, b2 = 1, b3 = 1)
sts1<-c(c1 = 1, c2 = 1, c3 = 1)
stt1<-c(Asym = 1, xmid = 1, scal = 1)
\end{verbatim}

@@

\begin{verbatim}
# Logistic3U
hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
  # This variant uses looping
  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt  <-  1:12
  res  <-  x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
}

hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
  jj  <-  matrix(0.0, 12, 3)
  tt  <-  1:12
  yy  <-  exp(-x[3] * tt)
  zz  <-  1.0 / (1 + x[2] * yy)
  jj[tt,1]   <-   zz
  jj[tt,2]   <-   -x[1] * zz * zz * yy
  jj[tt,3]   <-   x[1] * zz * zz * yy * x[2] * tt
  attr(jj, "gradient") <- jj
  jj
}
\end{verbatim}

@@

\begin{verbatim}
unls1<-try(nls(formula = frmu, start = stu1, data = weeddf))
\end{verbatim}

\begin{verbatim}
#> Error in nls(formula = frmu, start = stu1, data = weeddf) : 
#>   singular gradient
\end{verbatim}

\begin{verbatim}
snls1<-try(nls(formula = frms, start = sts1, data = weeddf))
\end{verbatim}

\begin{verbatim}
#> Error in nls(formula = frms, start = sts1, data = weeddf) : 
#>   singular gradient
\end{verbatim}

\begin{verbatim}
tnls1<-try(nls(formula = frmt, start = stt1, data = weeddf))
\end{verbatim}

\begin{verbatim}
#> Error in nls(formula = frmt, start = stt1, data = weeddf) : 
#>   singular gradient
\end{verbatim}

@@

\begin{verbatim}
unls1p<-try(nls(formula = frmu, start = stu1, data = weeddf, algorithm = "port"))
pnls(unls1p)
\end{verbatim}

\begin{verbatim}
#> unls1p  -- ss= 2.587277 :  b1 = 196.1863  b2 = 49.09164  b3 = 0.3135697; 17  itns
\end{verbatim}

\begin{verbatim}
snls1p<-try(nls(formula = frms, start = sts1, data = weeddf, algorithm = "port"))
pnls(snls1p)
\end{verbatim}

\begin{verbatim}
#> snls1p  -- ss= 2.587277 :  c1 = 1.961863  c2 = 4.909164  c3 = 3.135697; 15  itns
\end{verbatim}

\begin{verbatim}
tnls1p<-try(nls(formula = frmt, start = stt1, data = weeddf, algorithm = "port"))
pnls(tnls1p)
\end{verbatim}

\begin{verbatim}
#> tnls1p  -- ss= 2.587277 :  Asym = 196.1863  xmid = 12.4173  scal = 3.189083; 12  itns
\end{verbatim}

@@

\begin{verbatim}
unlx1 <- try(nlxb(formula = frmu, start = stu1, data = weeddf))
print(unlx1) 
\end{verbatim}

\begin{verbatim}
#> residual sumsquares =  2.5873  on  12 observations
#>     after  19    Jacobian and  25 function evaluations
#>   name            coeff          SE       tstat      pval      gradient    JSingval   
#> b1               196.186         11.31      17.35  3.167e-08  -4.859e-09        1011  
#> b2               49.0916         1.688      29.08  3.284e-10  -3.099e-08      0.4605  
#> b3               0.31357      0.006863      45.69  5.768e-12   2.305e-06     0.04714
\end{verbatim}

\begin{verbatim}
snlx1 <- try(nlxb(formula = frms, start = sts1, data = weeddf))
pshort(snlx1) # a short-form output
\end{verbatim}

\begin{verbatim}
#> snlx1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 34 res/ 23 jac
\end{verbatim}

\begin{verbatim}
tnlx1 <- try(nlxb(formula = frmt, start = stt1, data = weeddf))
pshort(tnlx1) # alternatively print(tnlx1)
\end{verbatim}

\begin{verbatim}
#> tnlx1  -- ss= 2.5873 :  Asym = 196.19  xmid = 12.417  scal = 3.1891; 36 res/ 27 jac
\end{verbatim}

@@

\begin{verbatim}
unlm1 <- try(nlsLM(formula = frmu, start = stu1, data = weeddf))
pnls(unlm1)  # Short form of output
\end{verbatim}

\begin{verbatim}
#> unlm1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
\end{verbatim}

\begin{verbatim}
snlm1 <- try(nlsLM(formula = frms, start = sts1, data = weeddf))
pnls(snlm1)
\end{verbatim}

\begin{verbatim}
#> snlm1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 7  itns
\end{verbatim}

\begin{verbatim}
tnlm1 <- try(nlsLM(formula = frmt, start = stt1, data = weeddf))
pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
\end{verbatim}

\begin{verbatim}
#> tnlm1  -- ss= 9205.4 :  Asym = 35.532  xmid = 43376  scal = -2935.4; 39  itns
\end{verbatim}

@@

\begin{verbatim}
library(gslnls)
ugslnls1 <- try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
pnls(ugslnls1) # to get sum of squares
\end{verbatim}

\begin{verbatim}
#> ugslnls1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
\end{verbatim}

\begin{verbatim}
sgslnls1 <- try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
pnls(sgslnls1) # Use summary() to get display
\end{verbatim}

\begin{verbatim}
#> sgslnls1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 9  itns
\end{verbatim}

\begin{verbatim}
tgslnls1 <- try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
pnls(tgslnls1) 
\end{verbatim}
